---
title: "BioMart Gene Ontology Gene Sets Collections"
author: "Zuguang Gu ( z.gu@dkfz.de )"
date: '`r Sys.Date()`'
output: html_document
---

<style type="text/css">
.main-container {
  max-width: 1000px;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r, echo = FALSE, message = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    echo = FALSE)
library(GetoptLong)
library(rGREAT)
```

First load the **BioMartGOGeneSets** package.

```{r, echo = TRUE}
library(BioMartGOGeneSets)
```

Get genes for a specific organism. The proper value of `dataset` can be found in the tables at the end of this document.

```{r, eval = FALSE, echo = TRUE}
getBioMartGenes(dataset)
```

Get GO gene sets for a specific organism.


```{r, eval = FALSE, echo = TRUE}
getBioMartGOGeneSets(dataset, ontology = "BP")
```


```{r}
table = readRDS("all_supported_organisms.rds")
table$genesets = paste0("BP (", table$n_bp_genesets, "), CC (", table$n_cc_genesets, "), MF (", table$n_mf_genesets, ")")
colnames(table)[colnames(table) == "n_gene"] = "genes"
df_genes = table[table$mart == "genes_mart", c(1:3, 5, 9), drop = FALSE]
df_plants = table[table$mart == "plants_mart", c(1:3, 5, 9), drop = FALSE]
df_metazoa = table[table$mart == "metazoa_mart", c(1:3, 5, 9), drop = FALSE]
df_fungi = table[table$mart == "fungi_mart", c(1:3, 5, 9), drop = FALSE]
df_protists = table[table$mart == "protists_mart", c(1:3, 5, 9), drop = FALSE]
```

Version information:

```{r, echo = TRUE}
BioMartGOGeneSets
```

Source code for retrieving GO gene sets can be obtained here:

```{r, echo = TRUE}
system.file("scripts", "biomart_genesets.R", package = "BioMartGOGeneSets")
```


<hr>

# {.tabset}

## Ensembl organisms (`r nrow(df_genes)` organisms)

```{r}
kable(df_genes, row.names = FALSE)
```

## Plants (`r nrow(df_plants)` organisms)

```{r}
kable(df_plants, row.names = FALSE)
```

## Metazoa (`r nrow(df_metazoa)` organisms)

```{r}
kable(df_metazoa, row.names = FALSE)
```

## Fungi (`r nrow(df_fungi)` organisms)

```{r}
kable(df_fungi, row.names = FALSE)
```

## Protists (`r nrow(df_protists)` organisms)

```{r}
kable(df_protists, row.names = FALSE)
```
